---
title: "PulsePAT-Nanostring"
author: "Thomas W. Battaglia"
output:
  pdf_document: default
  html_document:
    highlight: pygments
    theme: journal
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

### 1. Introduction
This is an Rmarkdown document which contains the R code for the Nanostring RNA-array analysis found within the manuscript "Ruiz et al. (2017)". Details about the tools and steps for processing the ileal transcriptome dataset can be found within the **'Methods'** section of the respective manuscript. The files used to generate this data set is hosted on GEO under the accession number [GSE98022](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE98022). This workflow will use a saved RDA file from the analysis instead of hosting the raw data.


#### 1a. Install and load the necessary libraries
```{r library, message = FALSE, warnings = FALSE}
source("scripts/geneExpress.R")
library(Biobase)
library(DESeq)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
library(stats)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
library(org.Mm.eg.db)
```


#### 1b. Import the eSet object containing the counts and metadata
```{r deseq}
load("data/nanostring_eset.rda")
```

----

### Compare Control vs. PAT1 vs. PAT3 at day 52 of life
```{r stats}
# Run significance testing function
## Output is an ExpressionSet, data.frame and significant-genes only data.frame
pulsePAT_results <- gene_express(expressionObject = pulsepat_eset)
```


### Figure 2a. Heatmap of significant genes between 3 groups
```{r heatmap, fig.height=15, fig.width=10}
# Subset genes by custom significance and gather gene names
pulsePAT_results_filtered = subset(pulsePAT_results$sig_results, 
                                   PAT1_Control_padj < 0.05 | PAT3_Control_padj < 0.05) %>%
  row.names()

# Subset output expresion object by only the significant genes
## make a matrix of values across all samples
mat <- pulsePAT_results$sig_expression[pulsePAT_results_filtered] %>%
  exprs()

# Choose which column variables you want to annotate the columns by.
annotation_col = data.frame(
  Group = factor(pulsePAT_results$sig_expression$Group), 
  row.names = colnames(pulsePAT_results$sig_expression)
)

# Specify colors you want to annotate the columns by.
ann_colors = list(
  Group = c(Control = "black", PAT3 = "deeppink", PAT1 = "green3")
)

# Plot heatmap with annotations
pheatmap(mat = mat, 
         color = colorRampPalette(c("#0000ff", "#000000", "#ffff00"))(n = 255), 
         scale = "row", 
         show_rownames = F,
         annotation_col = annotation_col, 
         annotation_colors = ann_colors,
         fontsize = 6.5,
         cellwidth = 8,
         cellheight = 2.5,
         show_colnames = F)
```

```{r}
sessionInfo()
```

