---
title: "PulsePAT-Nanostring"
author: "Thomas W. Battaglia"
output:
   pdf_document: 
    highlight: tango
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

### 1. Introduction
This is an Rmarkdown document which contains the R code for the Nanostring RNA-array analysis found within the manuscript "Ruiz et al. (2017)". Details about the tools and steps for processing the ileal transcriptome dataset can be found within the **'Methods'** section of the respective manuscript. The files used to generate this data set is hosted on GEO under the accession number [GSE98022](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE98022). This workflow will use a saved RDA file from the analysis instead of hosting the raw data.


#### 1a. Install and load the necessary libraries
```{r library, message = FALSE, warnings = FALSE}
source("scripts/geneExpress.R")
library(Biobase)
library(DESeq)
library(ggplot2)
library(tidyverse)
library(RColorBrewer)
library(stats)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
library(org.Mm.eg.db)
library(annotables)
```


#### 1b. Import the eSet object containing the counts and metadata
```{r deseq}
load("data/nanostring_eset.rda")
```

----

### Compare Control vs. PAT1 vs. PAT3 at day 52 of life
```{r stats}
# Run significance testing function
## Output is an ExpressionSet, data.frame and significant-genes only data.frame
pulsePAT_results <- gene_express(expressionObject = pulsepat_eset)

# Add genes as a rowname
pulsePAT_results$sig_results$gene_name = row.names(pulsePAT_results$sig_results)

# Subset genes by custom significance and gather gene names
pulsePAT_results_p1_p3 = subset(pulsePAT_results$sig_results, 
                                PAT1_Control_padj < 0.05 | PAT3_Control_padj < 0.05)

# Get the number of significant genes in both P1 + P3
nrow(pulsePAT_results_p1_p3)

# Get the number of genes significant in P1
pulsePAT_results_p1 = subset(pulsePAT_results$sig_results, PAT1_Control_padj < 0.05)

# Get the number of genes significant in P3
pulsePAT_results_p3 = subset(pulsePAT_results$sig_results, PAT3_Control_padj < 0.05)

# Find overlapping genes
dplyr::intersect(pulsePAT_results_p1$gene_name, pulsePAT_results_p3$gene_name) %>% length()

# Find P1-unique
dplyr::setdiff(pulsePAT_results_p1$gene_name, pulsePAT_results_p3$gene_name) %>% length()

# Find P3-unique
dplyr::setdiff(pulsePAT_results_p3$gene_name, pulsePAT_results_p1$gene_name) %>% length()
```

### Figure 2a. Heatmap of significant genes between 3 groups
```{r heatmap, fig.height=15, fig.width=10}
# Subset output expresion object by only the significant genes
## make a matrix of values across all samples
mat <- pulsePAT_results$sig_expression[row.names(pulsePAT_results_p1_p3)] %>%
  exprs()

# Choose which column variables you want to annotate the columns by.
annotation_col = data.frame(
  Group = factor(pulsePAT_results$sig_expression$Group), 
  row.names = colnames(pulsePAT_results$sig_expression)
)

# Specify colors you want to annotate the columns by.
ann_colors = list(
  Group = c(Control = "black", PAT1 = "#1E824C", PAT3 = "#ff0099")
)

# Plot heatmap with annotations
pheatmap(mat = mat, 
         color = colorRampPalette(c("#0000ff", "#000000", "#ffff00"))(n = 255), 
         scale = "row", 
         show_rownames = F,
         annotation_col = annotation_col, 
         annotation_colors = ann_colors,
         fontsize = 6.5, 
         border_color = NA,
         treeheight_row = 20,
         treeheight_col = 20,
         filename = "Figure2a.pdf",
         cellwidth = 8,
         cellheight = 2.5,
         show_colnames = F)
```

### Figure S3. Table of significant genes between 
```{r s6_table}
# Generate a .csv file for the significant genes
table_s3 = pulsePAT_results_p1_p3 %>%
  rownames_to_column(var = "geneid")

# Add genes names to the table
table_s3 = table_s3 %>%
  inner_join(grcm38, by = c("geneid" = "symbol"))

# Write file to disk  
write.csv(table_s3, "TableS3.csv", quote = F, row.names = F)
```


```{r}
sessionInfo()
```

