---
title: "PulsePAT-16S OTU"
author: "Thomas W. Battaglia"
output: 
  pdf_document: 
    highlight: tango
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

### Introduction
This Notebook is meant to contain the 16S_OTU data found within the respective manuscript. It contains only the code used to generate the figures found within the main text. This entire dataset is publically available in QIITA under the ID [10527](https://qiita.ucsd.edu/study/description/10527). More details about the procedures used to generate the data can be found within the **Methods** section of the manuscript. The table found within the data folder has been processed to remove any OTU less than 0.01% relative abundance.


#### 1a. Install and load the necessary libraries
```{r library, message = FALSE, warnings = FALSE}
source("scripts/estimate_pd.R")
source("scripts/intergroup_dist.R")
library(phyloseq)
library(RColorBrewer)
library(tidyverse)
library(vegan)
library(ape)
library(picante)
library(forcats)
```


#### 1b. Import the OTU table, map and OTU-representative tree
```{r import}
data <- import_biom(BIOMfilename = 'data/pulsePAT_m0001.biom', 
                    treefilename = 'data/rep_set.tre', 
                    parseFunction = parse_taxonomy_greengenes)
mapping = import_qiime_sample_data(mapfilename = 'data/pulsePAT_mapping.txt')
phylo <- merge_phyloseq(data, mapping)

# Convert time to factor
sample_data(phylo)$Day_of_life <- factor(sample_data(phylo)$Day_of_life)
```

----

#### Calculate Bdiv pairwise comparisons (Figure 1d)
```{r bdiv}
# Calculate bdiv intra group distances for pups
## Use function within Rscript
pups = phylo %>%
  subset_samples(Breeder_Pup == "pup") %>%
  subset_samples(Type_of_sample == "Fecal") %>%
  intergroup_div(class = "Treatment", timepoint = "Day_of_life", metric = "unifrac", ntimes = 999)

# Modify to calculate Mean/SEM means
pups_means = pups %>%
  mutate(MouseType = "Pups") %>%
  mutate(MouseType = factor(MouseType)) %>%
  filter(Comparison %in% c("ControlvsPAT1", "PAT3vsControl", "PAT1vsPAT3")) %>%
  mutate(Comparison = gsub("PAT3vsControl","ControlvsPAT3", Comparison, fixed = T)) %>%
  group_by(Comparison, timepoint, MouseType) %>%
  summarise(Mean_o_Mean = mean(Mean),
            sem_o_sem = mean(SEM)) %>%
  ungroup()

# Calculate bdiv intra group distances for pups
## Use function within Rscript
dams = phylo %>%
  subset_samples(Breeder_Pup == "Breeder") %>%
  subset_samples(Type_of_sample == "Fecal") %>%
  intergroup_div(class = "Treatment", timepoint = "Day_of_life", metric = "unifrac", ntimes = 999)

# Modify to calculate Mean/SEM means
dams_means = dams %>%
  mutate(MouseType = "Dams") %>%
  mutate(MouseType = factor(MouseType)) %>%
  filter(Comparison %in% c("ControlvsPAT")) %>%
  group_by(Comparison, timepoint, MouseType) %>%
  summarise(Mean_o_Mean = mean(Mean),
            sem_o_sem = mean(SEM)) %>%
  ungroup()

# Find uneven factors
pups_means$timepoint = as.character(pups_means$timepoint)
dams_means$timepoint = as.character(dams_means$timepoint)
pups_means$Comparison = as.character(pups_means$Comparison)
dams_means$Comparison = as.character(dams_means$Comparison)
pups_means$MouseType = as.character(pups_means$MouseType)
dams_means$MouseType = as.character(dams_means$MouseType)

# Combine two results
merged = bind_rows(pups_means, dams_means)

# Plot results over time
merged %>%
  mutate(Comparison = factor(Comparison),
         MouseType = factor(MouseType)) %>%
  mutate(timepoint = factor(timepoint, levels = c(5, 10, 12, 21, 22, 30, 34, 39, 43, 52))) %>%
  filter(timepoint %in% c(5, 10, 22, 30, 34, 39, 43, 52)) %>%
  filter(Comparison != "PAT1vsPAT3") %>%
  mutate(Comparison = gsub("vs", " vs ", Comparison, fixed = T)) %>%
  ggplot(aes(x = timepoint, y = Mean_o_Mean, group = Comparison, color = Comparison)) +
  geom_line(size = 1.20) +
  geom_point() +
  geom_errorbar(aes(ymin = Mean_o_Mean - sem_o_sem, ymax = Mean_o_Mean + sem_o_sem), width = 0.25) +
  annotate("rect", xmin = 1, xmax = 2, ymin = -Inf, 
           ymax = Inf, alpha = 0.2, fill = "darkgray", 
           color = "black", linetype = "dotted") +
  theme_bw(base_size = 15) +
  scale_color_manual(values = c("Control vs PAT1" = "#1E824C",
                                "Control vs PAT3" = "#ff0099",
                                "Control vs PAT" = "orange"), 
                     labels = c("Control vs PAT1" = "Control vs PAT1 (Pups)",
                                "Control vs PAT3" = "Control vs PAT3 (Pups)",
                                "Control vs PAT" = "Control vs PAT1 (Dams)")) +
  xlab("Day of life (pups)") +
  theme(legend.title = element_blank(), 
        legend.position = c(0.80, 0.15),
        legend.text = element_text(size = 16),
        legend.background = element_rect(size = 0.55, linetype = "solid", colour = "gray")) +
  ylab("Mean Pairwise Unweighted \nUniFrac Distances (Inter-group)") +
  ggsave('Figure1d.pdf')
```

----

#### Calculate Bdiv pairwise comparisons (Figure S1-b)
```{r figuresS1_b}
# Calculate bdiv intra group distances for pups
## Use function within Rscript
pups_all = phylo %>%
  subset_samples(Breeder_Pup == "pup") %>%
  subset_samples(Type_of_sample == "Fecal") %>%
  intergroup_div(class = "Treatment", timepoint = "Day_of_life", metric = "unifrac", ntimes = 9) %>%
  mutate(MouseType = "Pups") %>%
  group_by(Comparison, timepoint, MouseType) %>%
  summarise(Mean_o_Mean = mean(Mean),
            sem_o_sem = mean(SEM)) %>%
  ungroup() %>%
  mutate(timepoint = as.character(timepoint)) %>%
  mutate(Comparison = as.character(Comparison)) 

# Calculate bdiv intra group distances for dams
## Use function within Rscript
dams_all = phylo %>%
  subset_samples(Breeder_Pup == "Breeder") %>%
  subset_samples(Type_of_sample == "Fecal") %>%
  intergroup_div(class = "Treatment", timepoint = "Day_of_life", metric = "unifrac", ntimes = 9) %>%
  mutate(MouseType = "Dams") %>%
  mutate(Comparison = gsub("ControlvsControl", "ControlvsControl_D", Comparison, fixed = T)) %>%
  group_by(Comparison, timepoint, MouseType) %>%
  summarise(Mean_o_Mean = mean(Mean),
            sem_o_sem = mean(SEM)) %>%
  ungroup() %>%
  mutate(timepoint = as.character(timepoint)) %>%
  mutate(Comparison = as.character(Comparison)) 

# Merge tables together
merged_all = bind_rows(pups_all, dams_all)

# Plot multi-dimensional barplot
merged_all %>%
  filter(Comparison != "PAT1vsPAT3") %>%
  mutate(Comparison = factor(Comparison)) %>%
  mutate(MouseType = factor(MouseType)) %>%
  filter(timepoint %in% c(5, 10, 22, 39, 43, 52)) %>%
  mutate(timepoint = factor(timepoint, levels = c(5, 10, 22, 39, 43, 52))) %>%
  mutate(Com_labels = case_when(.$Comparison == "PAT3vsPAT3" ~ "P3 vs P3",
                                .$Comparison == "PAT3vsControl" ~ "C vs P3",
                                .$Comparison == "ControlvsControl" ~ "C vs C",
                                .$Comparison == "ControlvsPAT1" ~ "C vs P1",
                                .$Comparison == "PAT1vsPAT1" ~ "P1 vs P1",
                                .$Comparison == "PATvsPAT" ~ "P vs P",
                                .$Comparison == "ControlvsPAT" ~ "C vs P",
                                .$Comparison == "ControlvsControl_D" ~ "C vs CD")) %>%
   mutate(Com_labels = factor(Com_labels, levels = c("P3 vs P3", "C vs P3", "C vs C", "C vs P1", "P1 vs P1", 
                                                   "P vs P", "C vs P", "C vs CD"))) %>%
  ggplot(aes(x = Com_labels, y = Mean_o_Mean, fill = Com_labels, color = Com_labels)) +
  geom_bar(stat = "identity", size = 0.70) +
  coord_flip() +
  facet_wrap(~ timepoint, ncol = 2, scale = "free_y") + 
  xlab("") + ylab("Mean unweighted UniFrac distances") +
  theme_bw(base_size = 14) +
  geom_errorbar(aes(ymin = Mean_o_Mean, ymax = Mean_o_Mean + sem_o_sem)) +
  scale_fill_manual(name = "",
                    values = c("P3 vs P3" = "#DA3C78", 
                               "C vs P3" = "#DA3C78", 
                               "C vs C" = "#7f8c8d",
                               "C vs P1" = "#26A65B",
                               "P1 vs P1" = "#26A65B",
                               "P vs P" = "#F1892D",
                               "C vs P" = "#F1892D",
                               "C vs CD" = "#7f8c8d")) +
  scale_color_manual(name = "",
                     values = c("P3 vs P3" = "#DA3C78", 
                               "C vs P3" = "black", 
                               "C vs C" = "#7f8c8d",
                               "C vs P1" = "black",
                               "P1 vs P1" = "#26A65B",
                               "P vs P" = "#F1892D",
                               "C vs P" = "black",
                               "C vs CD" = "#7f8c8d")) +
  ggsave("FigureS1-b.pdf", width = 9, height = 11)
```




```{r}
sessionInfo()
```