---
title: "GF-PAT-Nanostring"
author: "Thomas W. Battaglia"
output:
  pdf_document: default
  html_document:
    highlight: pygments
    theme: journal
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

### 1. Introduction
This is an Rmarkdown document which contains the R code for the Nanostring RNA-array analysis found within the manuscript "Ruiz et al. (2017)". Details about the tools and steps for processing the ileal transcriptome dataset can be found within the **'Methods'** section of the respective manuscript. The files used to generate this data set is hosted on GEO under the accession number [GSE98291](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE98291). This workflow will use a saved RDA file from the analysis instead of hosting the raw data.


#### 1a. Install and load the necessary libraries
```{r library, message = FALSE, warnings = FALSE}
source("scripts/geneExpress.R")
library(Biobase)
library(DESeq)
library(ggplot2)
library(tidyverse)
library(RColorBrewer)
library(stats)
library(ggbiplot)
library(pheatmap)
library(org.Mm.eg.db)
library(annotables)
```


#### 1b. Import the eSet object containing the counts and metadata
```{r deseq}
load("data/nanostring_eset.rda")
```

----

### Compare Control-SPF vs. PAT-SPF vs. Control-GF vs. PAT-GF at day 52 of life
```{r stats}
# Run significance testing function
## Output is an ExpressionSet, data.frame and significant-genes only data.frame
gfPAT_results <- gene_express(expressionObject = gfpat_eset)

# Subset genes by custom significance and gather gene names
gf_results_filtered = subset(gfPAT_results$sig_results, 
                             PATSPF_ControlSPF_padj < 0.05 | ControlSPF_ControlGF_padj < 0.05) 

# Get the number of significant genes in both P1 + P3
nrow(gf_results_filtered)

# Get the number of genes significant
gfPAT_results_spf_trt = subset(gfPAT_results$sig_results, PATSPF_ControlSPF_padj < 0.05)

# Get the number of genes significant
gfPAT_results_ctl = subset(gfPAT_results$sig_results, ControlSPF_ControlGF_padj < 0.05)

# Get the number of genes significant
gfPAT_results_gf = subset(gfPAT_results$sig_results, PATGF_ControlGF_padj < 0.05)
```


### Figure 3g. Heatmap of significant genes between 3 groups
```{r heatmap, fig.height=10, fig.width=10}
# Subset output expresion object by only the significant genes
## make a matrix of values across all samples
mat <- gfPAT_results$sig_expression[row.names(gf_results_filtered)] %>%
  exprs()

# Make consistent grouping names across figures
Groups <- factor(gfPAT_results$sig_expression$Group, 
                 labels = c("GF: Control", "SPF: Control","GF: PAT", "SPF: PAT"))

# Choose which column variables you want to annotate the columns by.
annotation_col = data.frame(
  Group = Groups, 
  row.names = colnames(gfPAT_results$sig_expression)
)

# Specify colors you want to annotate the columns by.
ann_colors = list(
  Group = c("SPF: Control" = "darkblue", 
            "SPF: PAT" = "red", 
            "GF: Control" = "gray", 
            "GF: PAT" = "orange"))

# Plot heatmap with annotations (Figure 3g)
pheatmap(mat = mat, 
         color = colorRampPalette(c("#0000ff", "#000000", "#ffff00"))(n = 100), 
         scale = "row", 
         show_rownames = F,
         annotation_col = annotation_col, 
         annotation_colors = ann_colors, 
         filename = "Figure3g.pdf", 
         border_color = NA,
         treeheight_row = 20,
         treeheight_col = 20,
         fontsize = 6.5,
         cellwidth = 20,
         show_colnames = F)
```


### Figure 3f. PCoA Biplot for all 3 groups
```{r biplot, fig.height=15, fig.width=10}
# Generate Principal Components Analysis data for normalized-log2 values
pca <- gfPAT_results$sig_expression %>%
  exprs() %>%
  t() %>%
  prcomp(scale. = T)

# Use ggbiplot to create a biplot
ggbiplot(pca, 
         obs.scale = 1, 
         var.scale = 1, 
         groups = as.vector(gfPAT_results$sig_expression$Group), 
         ellipse = TRUE, 
         circle = TRUE, 
         var.axes = F) +
  scale_color_manual(breaks = c("ControlSPF", "PATSPF", "ControlGF", "PATGF"), 
                     labels = c("SPF: Control", "SPF: PAT", "GF: Control", "GF: PAT"),
                     values = c("ControlSPF" = "darkblue", 
                                "PATSPF" = "red", 
                                "ControlGF" = "grey", 
                                "PATGF" = "orange")) + 
  theme_bw(base_size = 17) +
  geom_point(aes(color = factor(gfPAT_results$sig_expression$Group)), size = 4, alpha = 0.65) +
  theme(legend.title = element_blank(), legend.position = c(0.89, 0.16),
        legend.background = element_rect(size = 0.5, linetype= "solid", colour = "gray")) +
  ggsave("Figure3f.pdf", width = 10, height = 8)
```


### Figure S6. Table of significant genes between GF-PAT vs. GF-Control
```{r s6_table}
# Generate a .csv file for the significant genes
table_s6 = subset(gfPAT_results$sig_results, 
                  PATSPF_ControlSPF_padj < 0.05 | ControlSPF_ControlGF_padj < 0.05 | PATGF_ControlGF_padj < 0.05) %>%
  rownames_to_column(var = "geneid")

# Add genes names to the table
table_s6 = table_s6 %>%
  inner_join(grcm38, by = c("geneid" = "symbol"))

# Write file to disk  
write.csv(table_s6, "TableS6.csv", quote = F, row.names = F)
```




```{r}
sessionInfo()
```

